---
title: "Analyzing Music Data"
author: "Hyder Ali"
format: html
execute: 
  echo: false
  message: false
  warning: false
  cache: true
---

```{r}
library(tidyverse)
```

## Introduction & Dataset Preview

We are analyzing Billboard Hot 100 chart data from the year 2000. Below is a preview of the first 10 rows of the dataset, showing the artist, track, entry date, and the weekly rankings for the first four weeks:

```{r}
billboard |>
  select(artist, track, date.entered, wk1, wk2, wk3, wk4) |>
  head(10) |>
  knitr::kable()
```

## Analysis of Entry Dates

Below are the summary statistics for when songs entered the Billboard Hot 100 chart:

```{r}
summary(billboard$date.entered)
```


## Distribution of Chart Ranks

Next, we analyze the distribution of chart ranks at two different points in time: during the first week of entry (Week 1) and after six weeks on the chart (Week 6).

### First-Week Chart Rank Distribution

This chart illustrates where songs typically debut on the Hot 100. Lower values indicate higher (better) chart rankings.

```{r}
wk1_count <- billboard |>
  summarise(n = sum(!is.na(wk1))) |>
  pull(n)

billboard |>
  filter(!is.na(wk1)) |>
  ggplot(aes(x = wk1, fill = "Initial Entry (Week 1)")) +
  geom_histogram(binwidth = 2, color = "black", alpha = 0.85) +
  scale_fill_manual(
    values = c("Initial Entry (Week 1)" = "#2b5c8f")
  ) +
  scale_x_continuous(breaks = seq(0, 100, 10), limits = c(1, 100)) +
  labs(
    title = "Billboard Hot 100: Initial Entry Position Distribution",
    subtitle = paste0("First-week chart positions (wk1) for debuting songs (n = ", wk1_count, ")"),
    x = "Chart Position (1 = Best, 100 = Worst)",
    y = "Number of Songs",
    fill = "Chart Phase"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, color = "#2c3e50"),
    plot.subtitle = element_text(size = 10, color = "#7f8c8d", margin = margin(b = 10)),
    axis.title = element_text(face = "bold", color = "#34495e"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
```

### Sixth-Week Chart Rank Distribution

By the sixth week, some songs have dropped off the chart, while others have moved up. This chart shows the distribution of ranks for the remaining active songs.

```{r}
wk6_count <- billboard |>
  summarise(n = sum(!is.na(wk6))) |>
  pull(n)

billboard |>
  filter(!is.na(wk6)) |>
  ggplot(aes(x = wk6, fill = "Active Run (Week 6)")) +
  geom_histogram(binwidth = 2, color = "black", alpha = 0.85) +
  scale_fill_manual(
    values = c("Active Run (Week 6)" = "#e65c00")
  ) +
  scale_x_continuous(breaks = seq(0, 100, 10), limits = c(1, 100)) +
  labs(
    title = "Billboard Hot 100: Six-Week Progression Position Distribution",
    subtitle = paste0("Chart positions after six weeks (wk6) on the Hot 100 (n = ", wk6_count, ")"),
    x = "Chart Position (1 = Best, 100 = Worst)",
    y = "Number of Songs",
    fill = "Chart Phase"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, color = "#2c3e50"),
    plot.subtitle = element_text(size = 10, color = "#7f8c8d", margin = margin(b = 10)),
    axis.title = element_text(face = "bold", color = "#34495e"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
```

## Missing and Present Rankings Analysis

To see how many songs remain on the chart over time, we calculate the number of present (active) and missing (dropped off) rankings for selected weeks:

```{r}
billboard |>
  select(wk1, wk4, wk10, wk20, wk40, wk76) |>
  pivot_longer(everything(), names_to = "week", values_to = "rank") |>
  group_by(week) |>
  summarise(
    Present = sum(!is.na(rank)),
    Missing = sum(is.na(rank))
  ) |>
  mutate(week = factor(week, levels = c("wk1", "wk4", "wk10", "wk20", "wk40", "wk76"))) |>
  arrange(week) |>
  rename(Week = week) |>
  knitr::kable()
```

## Song Chart Re-entry Analysis

We investigate whether any songs leave the Hot 100 chart and later re-enter it. A song is considered to have re-entered if there is a gap of one or more weeks (recorded as `NA`) between active chart positions:

```{r}
billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank"
  ) |>
  mutate(week_num = as.integer(str_extract(week, "\\d+"))) |>
  filter(!is.na(rank)) |>
  group_by(artist, track) |>
  arrange(week_num) |>
  mutate(week_diff = week_num - lag(week_num)) |>
  summarise(re_entered = any(week_diff > 1, na.rm = TRUE), .groups = "drop") |>
  count(re_entered) |>
  mutate(
    re_entered = if_else(re_entered, "Re-entered Chart", "Never Re-entered")
  ) |>
  rename(`Song Status` = re_entered, `Number of Songs` = n) |>
  knitr::kable()
```

## Week 1 vs. Week 6 Rank Comparison

We compare each song's Week 1 (debut) rank and Week 6 rank. We categorize songs based on whether their chart rank improved (rank number decreased), stayed the same, got worse (rank number increased), or if they were missing (dropped off the chart) by Week 6:

```{r}
# 1. Compare ranks and count status
status_table <- billboard |>
  mutate(
    status = case_when(
      is.na(wk6) ~ "Missing by Week 6",
      wk6 < wk1  ~ "Improved",
      wk6 == wk1 ~ "Stayed the Same",
      wk6 > wk1  ~ "Got Worse"
    )
  ) |>
  count(status) |>
  rename(`Week 6 Status` = status, `Number of Songs` = n)

knitr::kable(status_table)
```

For songs that remained on the chart in both Week 1 and Week 6, we calculate the median change in rank. Note that positive numbers for positions climbed represent improvement (e.g., climbing from rank 80 to rank 53 is +27 positions):

```{r}
median_metrics <- billboard |>
  filter(!is.na(wk1) & !is.na(wk6)) |>
  mutate(
    change_climbed = wk1 - wk6,
    change_numerical = wk6 - wk1
  ) |>
  summarise(
    `Median Positions Climbed (wk1 - wk6)` = median(change_climbed),
    `Median Numerical Rank Change (wk6 - wk1)` = median(change_numerical)
  )

knitr::kable(median_metrics)
```

## Song Rankings Over Time

We reshape the `billboard` dataset from wide format to long format, converting all the week columns (`wk1` to `wk76`) into two columns: `week` and `rank`. We then plot the trajectory of each song's ranking over time:

```{r}
billboard_long <- billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) |>
  mutate(
    week = as.integer(str_extract(week, "\\d+"))
  )

billboard_long |>
  ggplot(aes(x = week, y = rank, group = interaction(artist, track))) +
  geom_line(alpha = 0.15, color = "#2c3e50") +
  scale_y_reverse(breaks = seq(1, 100, 10), limits = c(100, 1)) +
  scale_x_continuous(breaks = seq(1, 76, 5)) +
  labs(
    title = "Song Rank Trajectories Over Time",
    subtitle = "Individual trajectories for all 317 songs on the Billboard Hot 100",
    x = "Weeks on Chart",
    y = "Chart Rank (1 = Best, 100 = Worst)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, color = "#2c3e50"),
    plot.subtitle = element_text(size = 10, color = "#7f8c8d", margin = margin(b = 10)),
    axis.title = element_text(face = "bold", color = "#34495e")
  )
```

## Song-Level Summary and Notable Achievements

We aggregate the reshaped billboard dataset to create a song-level summary. For each track, we identify its debut (first) rank, its best (peak) rank, the first week it reached its peak rank, and the total number of weeks it spent active on the Hot 100 chart.

Here is a preview of the first 10 songs in this summary:

```{r}
# 1. Generate song-level summary
song_summary <- billboard_long |>
  group_by(artist, track) |>
  summarise(
    first_rank = rank[which.min(week)],
    best_rank = min(rank),
    first_week_at_best = min(week[rank == best_rank]),
    total_weeks = n(),
    .groups = "drop"
  )

song_summary |>
  head(10) |>
  rename(
    Artist = artist,
    Track = track,
    `First Rank` = first_rank,
    `Best Rank` = best_rank,
    `Week of Best Rank` = first_week_at_best,
    `Total Weeks` = total_weeks
  ) |>
  knitr::kable()
```

### Notable Top-10 Song Achievements

Based on this summary, we identify three notable achievements among songs that reached the Top-10:
* **The Fastest Song to Reach Number One:** The song that reached the #1 spot in the fewest weeks.
* **The Slowest Song to Reach Number One:** The song that reached the #1 spot in the most weeks.
* **The Top-10 Song with the Longest Chart Run:** The track that peaked in the top-10 and logged the most total weeks on the chart.

```{r}
# 2. Identify achievements
fastest_no1 <- song_summary |>
  filter(best_rank == 1) |>
  slice_min(first_week_at_best, n = 1) |>
  mutate(Category = "Fastest to reach #1")

slowest_no1 <- song_summary |>
  filter(best_rank == 1) |>
  slice_max(first_week_at_best, n = 1) |>
  mutate(Category = "Slowest to reach #1")

longest_run_top10 <- song_summary |>
  filter(best_rank <= 10) |>
  slice_max(total_weeks, n = 1) |>
  mutate(Category = "Longest chart run (Top-10)")

# Combine and print table
notable_songs <- bind_rows(fastest_no1, slowest_no1, longest_run_top10) |>
  select(Category, artist, track, first_rank, best_rank, first_week_at_best, total_weeks) |>
  rename(
    Artist = artist,
    Track = track,
    `First Rank` = first_rank,
    `Best Rank` = best_rank,
    `Week of Best Rank` = first_week_at_best,
    `Total Weeks` = total_weeks
  )

knitr::kable(notable_songs)
```

## Plotting Top-10 Song Trajectories

Below, we filter the reshaped dataset to include only songs that reached the Top-10 at some point during their run. We plot their trajectories in gray, and highlight our three notable songs in distinct colors to compare their journeys on the chart:

```{r}
# 1. Define the notable song labels for joining
notable_labels <- tibble(
  artist = c("Madonna", "Lonestar", "Creed"),
  track = c("Music", "Amazed", "Higher"),
  notable_name = c("Madonna - Music (Fastest to #1)", 
                   "Lonestar - Amazed (Slowest to #1)", 
                   "Creed - Higher (Longest Run)")
)

# 2. Reshape, filter for Top-10 songs, and set up highlight categories
top10_trajectories <- billboard_long |>
  group_by(artist, track) |>
  filter(min(rank) <= 10) |>
  ungroup() |>
  left_join(notable_labels, by = c("artist", "track")) |>
  mutate(
    highlight_group = ifelse(is.na(notable_name), "Other Top-10 Songs", notable_name),
    highlight_group = factor(
      highlight_group,
      levels = c(
        "Madonna - Music (Fastest to #1)",
        "Lonestar - Amazed (Slowest to #1)",
        "Creed - Higher (Longest Run)",
        "Other Top-10 Songs"
      )
    )
  ) |>
  arrange(desc(highlight_group)) # Ensures highlighted lines are plotted on top

# 3. Create the highlighted trajectory plot
top10_trajectories |>
  ggplot(aes(x = week, y = rank, group = interaction(artist, track), 
             color = highlight_group, alpha = highlight_group, linewidth = highlight_group)) +
  geom_line() +
  scale_y_reverse(breaks = seq(1, 100, 10), limits = c(100, 1)) +
  scale_x_continuous(breaks = seq(1, 76, 5)) +
  scale_color_manual(
    values = c(
      "Madonna - Music (Fastest to #1)" = "#d95f02",
      "Lonestar - Amazed (Slowest to #1)" = "#7570b3",
      "Creed - Higher (Longest Run)" = "#1b7837",
      "Other Top-10 Songs" = "#d3d3d3"
    )
  ) +
  scale_linewidth_manual(
    values = c(
      "Madonna - Music (Fastest to #1)" = 1.1,
      "Lonestar - Amazed (Slowest to #1)" = 1.1,
      "Creed - Higher (Longest Run)" = 1.1,
      "Other Top-10 Songs" = 0.4
    )
  ) +
  scale_alpha_manual(
    values = c(
      "Madonna - Music (Fastest to #1)" = 1.0,
      "Lonestar - Amazed (Slowest to #1)" = 1.0,
      "Creed - Higher (Longest Run)" = 1.0,
      "Other Top-10 Songs" = 0.35
    )
  ) +
  labs(
    title = "Top-10 Song Rank Trajectories",
    subtitle = "Highlighting the fastest/slowest number-one songs and the longest running top-10 song",
    x = "Weeks on Chart",
    y = "Chart Rank (1 = Best, 100 = Worst)",
    color = "Song Highlight",
    linewidth = "Song Highlight",
    alpha = "Song Highlight"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, color = "#2c3e50"),
    plot.subtitle = element_text(size = 10, color = "#7f8c8d", margin = margin(b = 10)),
    axis.title = element_text(face = "bold", color = "#34495e"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
```
